---
title: "Financial Literacy Analysis"
author: "Mariana Garcia Mejia"
output:
  html_document: default
  pdf_document: default
---

```{r message=FALSE, warning=FALSE}
# LOAD LIBRARIES
library(tidyverse)   # dplyr, tidyr, ggplot2
library(scales)      # percent formatting on axes

# Regression libraries
library(car)         # vif()
library(lmtest)      # dwtest()
library(stargazer)   # regression tables
library(fastDummies) # dummy variables
```

```{r load-data}
# LOAD DATASET
df <- read.csv("analysis_data.csv", fileEncoding = "latin1")
```

## 1. Exploratory Data Analysis

```{r scores-summary}
# Summary of the scores
indices <- c("advanced", "basic", "overall")
averages <- c(mean(df$advanced_index), mean(df$basic_index), mean(df$overall_index))
sd <- c(sd(df$advanced_index), sd(df$basic_index), sd(df$overall_index))

summary <- tibble::tibble(indices, averages, sd)
summary
```

```{r gender-pie}
# Gender distribution — pie chart
gender_counts <- df %>%
  count(gender) %>%
  filter(!is.na(gender))

ggplot(gender_counts, aes(x = "", y = n, fill = gender)) +
  geom_col(width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(n / sum(n) * 100), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  labs(title = "Gender Distribution", fill = NULL)
```

```{r summary-gender}
# Stock market participation by gender
summary_gender <- df %>%
  filter(gender %in% c("male", "female", "non binary")) %>%
  group_by(gender) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    n_no               = n_obs - n_yes,
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  )

print(summary_gender)
```

```{r summary-undergrad}
# Stock market participation by undergraduate category
summary_undergrad <- df %>%
  group_by(undergrad_category) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    n_no               = n_obs - n_yes,
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  )

print(summary_undergrad)
```

```{r summary-country}
# Stock market participation by country of birth
summary_country <- df %>%
  group_by(born_in_canada) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  ) %>%
  mutate(born_in_canada = ifelse(born_in_canada == 1, "Born in Canada", "Born Outside Canada"))

print(summary_country)
```

```{r summary-scores-binary}
# Average literacy scores by stock market participation
summary_scores <- df %>%
  group_by(stock_mkt_participation_clean) %>%
  summarise(
    n_obs         = n(),
    avg_basic     = round(mean(basic_index,    na.rm = TRUE), 2),
    avg_advanced  = round(mean(advanced_index, na.rm = TRUE), 2),
    avg_overall   = round(mean(overall_index,  na.rm = TRUE), 2)
  ) %>%
  mutate(stock_mkt_participation_clean = ifelse(
    stock_mkt_participation_clean == 1,
    "Participates in Stock Market", "Does Not Participate / Refusal"
  ))

print(summary_scores)
```

# 2. Figures in the final paper

## Figure 1: Distribution of Index Scores
```{r figure1}
# Figure 1: Distribution of Index Scores
df %>%
  select(basic_index, advanced_index, overall_index) %>%
  pivot_longer(everything(), names_to = "Index", values_to = "Value") %>%
  mutate(Index = recode(Index,
                        "basic_index"    = "Basic Index",
                        "advanced_index" = "Advanced Index",
                        "overall_index"  = "Overall Index")) %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30, fill = "#0072B2", color = "white") +
  facet_wrap(~Index) +
  labs(
    title = "Distribution of Index Scores",
    x = "Value", y = "Frequency"
  ) +
  theme_minimal()
```

## Figure 2: Proportion of Correct Responses by Question Type
```{r figure2}
basic_questions    <- c("q1_basic_correct", "q2_basic_correct",
                        "q3_basic_correct", "q4_basic_correct")
advanced_questions <- c("q1_advanced_correct", "q2_advanced_correct", "q3_advanced_correct",
                        "q4_advanced_correct", "q5_advanced_correct")

basic_labels    <- c("Q1 — Numeracy", "Q2 — Inflation",
                     "Q3 — Time Value", "Q4 — Money Illusion")
advanced_labels <- c("Q1 — Stock Market Function", "Q2 — Stock Ownership",
                     "Q3 — Bond Ownership", "Q4 — Diversification",
                     "Q5 — Bond/Interest Rate")

basic_df <- data.frame(
  Question = basic_labels,
  Proportion = colMeans(df[, basic_questions], na.rm = TRUE),
  Type = "Basic"
)
advanced_df <- data.frame(
  Question = advanced_labels,
  Proportion = colMeans(df[, advanced_questions], na.rm = TRUE),
  Type = "Advanced"
)

all_q <- bind_rows(basic_df, advanced_df)

ggplot(all_q, aes(x = Question, y = Proportion)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~Type, scales = "free_x") +
  labs(
    title = "Proportion of Correct Responses by Question Type",
    x = NULL, y = "Proportion Correct"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Figure 3: Questions Breakdown


## Figure 4: Financial Literacy by Gender
```{r figure4}
df_gender <- df %>%
  filter(gender %in% c("male", "female")) %>%
  group_by(gender) %>%
  summarise(
    Basic    = mean(basic_index,    na.rm = TRUE),
    Advanced = mean(advanced_index, na.rm = TRUE),
    Overall  = mean(overall_index,  na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Basic, Advanced, Overall),
               names_to = "Index", values_to = "Score")

palette <- c("female" = "#FF7F0E", "male" = "#0072B2")

ggplot(df_gender, aes(x = Index, y = Score, fill = gender)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = palette, labels = c("Female", "Male")) +
  ylim(0, 1) +
  labs(
    title = "Financial Literacy by Gender",
    x = NULL, y = "Average Score", fill = NULL
  ) +
  theme_minimal()
```

## Figure 5: Stock Market Participation Pie Chart
```{r figure5}
stock_counts <- df %>%
  count(stock_mkt_participation) %>%
  filter(!is.na(stock_mkt_participation))

ggplot(stock_counts, aes(x = "", y = n, fill = stock_mkt_participation)) +
  geom_col(width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(n / sum(n) * 100), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  labs(
    title = "Do you own individual stocks, bonds, mutual funds or index funds?",
    fill = NULL
  )
```

## Figure 6: Stock Market Participation by Undergraduate Category
```{r}
# Figure 6: Stock Market Participation by Undergraduate Category
df %>%
  mutate(Participation = ifelse(stock_mkt_participation_clean == 1, "Yes", "No")) %>%
  ggplot(aes(x = undergrad_category, fill = Participation)) +
  geom_bar(position = "dodge") +
  labs(
    title  = "Stock Market Participation by Undergraduate Category",
    x      = "\nUndergraduate Category",
    y      = "Number of Respondents",
    fill   = "Stock Market Participation"
  ) +
  theme_minimal()
```

## Figure 7: Financial Literacy Scores by Stock Market Participation
```{r figure7}
means_long <- df %>%
  group_by(stock_mkt_participation_clean) %>%
  summarise(
    Basic    = mean(basic_index,    na.rm = TRUE),
    Advanced = mean(advanced_index, na.rm = TRUE),
    Overall  = mean(overall_index,  na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Basic, Advanced, Overall),
               names_to = "Index", values_to = "Score") %>%
  mutate(participation = ifelse(stock_mkt_participation_clean == 1, "Yes", "No / Refusal"))

ggplot(means_long, aes(x = participation, y = Score, fill = Index)) +
  geom_col(position = "dodge") +
  ylim(0, 0.8) +
  labs(
    title = "Financial Literacy Scores by Stock Market Participation",
    x = NULL, y = "Average Financial Literacy Score (%)", fill = NULL
  ) +
  theme_minimal()
```

# Regressions

```{r warning=FALSE}
# DATA PREPARATION

df <- df %>%
  mutate(
    stock_mkt_binary   = ifelse(trimws(tolower(stock_mkt_participation)) == "yes", 1L, 0L),
    gender_clean       = trimws(tolower(gender)),
    born_in_canada_bin = ifelse(trimws(tolower(as.character(born_in_canada)))
                                  %in% c("true", "yes", "1"), 1L, 0L)
  ) %>%
  filter(gender_clean %in% c("male", "female"))

df_model <- df %>%
  mutate(
    undergrad_category = trimws(tolower(undergrad_category)),
    gender_clean       = trimws(tolower(gender_clean))
  ) %>%
  fastDummies::dummy_cols(
    select_columns           = c("undergrad_category", "gender_clean"),
    remove_selected_columns  = TRUE,
    remove_most_frequent_dummy = FALSE
  )

# Fix all column names — replace spaces and special characters with underscores
names(df_model) <- gsub("[^a-zA-Z0-9_]", "_", names(df_model))

# Print column names to verify
# names(df_model)

# Identify and drop reference dummies
ref_undergrad <- names(df_model)[grepl("undergrad_category_", names(df_model)) &
                                   grepl("arts", names(df_model))]
ref_gender    <- names(df_model)[grepl("gender_clean_female", names(df_model))]

undergrad_dummies <- setdiff(
  names(df_model)[grepl("undergrad_category_", names(df_model))],
  ref_undergrad
)
gender_dummies <- setdiff(
  names(df_model)[grepl("gender_clean_", names(df_model))],
  ref_gender
)

# Assemble modelling data and drop NAs
model_vars <- c("age", "duration_minutes", "born_in_canada_bin",
                "stock_mkt_binary", "overall_index", "basic_index", "advanced_index",
                gender_dummies, undergrad_dummies)

predictors <- df_model %>%
  select(all_of(model_vars)) %>%
  drop_na() %>%
  mutate(across(everything(), as.numeric))

cat("Sample size after dropping missing values:", nrow(predictors), "\n")
cat("Stock market participants:", sum(predictors$stock_mkt_binary), "\n")
cat("Non-participants:", sum(predictors$stock_mkt_binary == 0), "\n")
```

```{r ols-regression}
# OLS: OVERALL ~ DEMOGRAPHICS

undergrad_dummies <- c(
  "undergrad_category_education___child_studies",
  "undergrad_category_humanities___languages",
  "undergrad_category_social_sciences",
  "undergrad_category_stem___health_sciences"
)

gender_dummies <- c("gender_clean_male")

lm_cols <- c("age", "duration_minutes", "born_in_canada_bin",
             gender_dummies, undergrad_dummies)

lm_model <- lm(
  reformulate(lm_cols, response = "overall_index"),
  data = predictors
)

summary(lm_model)
```

```{r ols-diagnostics, fig.width=12, fig.height=10}
# OLS ASSUMPTION DIAGNOSTICS
par(mfrow = c(2, 2))
plot(lm_model,
     sub.caption = "Figure 11: Linear Regression — Assumption Diagnostic Plots")

# Durbin-Watson
dw <- dwtest(lm_model)
cat(sprintf("\nDurbin-Watson statistic: %.3f\n", dw$statistic))
cat(sprintf("  p-value: %.4f\n", dw$p.value))
cat(sprintf("  → %s\n",
            ifelse(dw$statistic > 1.5 & dw$statistic < 2.5,
                   "OK — no autocorrelation concern", "Flag in paper")))
```

```{r warning=FALSE}
# LOGISTIC REGRESSION: FULL
glm_cols <- c("overall_index", "age", "duration_minutes", "born_in_canada_bin",
              gender_dummies, undergrad_dummies)

glm_model <- glm(
  reformulate(glm_cols, response = "stock_mkt_binary"),
  data   = predictors,
  family = binomial(link = "logit")
)

summary(glm_model)

or_full <- data.frame(
  Odds_Ratio = exp(coef(glm_model)),
  CI_Lower   = exp(confint(glm_model)[, 1]),
  CI_Upper   = exp(confint(glm_model)[, 2]),
  p_value    = coef(summary(glm_model))[, 4]
)

print(round(or_full, 3))
```

```{r logistic-assumptions}
# LOGISTIC REGRESSION: ASSUMPTIONS

# Sample size adequacy
n_events     <- sum(predictors$stock_mkt_binary)
n_predictors <- length(glm_cols)
cat(sprintf("\nSample size adequacy:\n  Events: %d | Predictors: %d | Min required: %d\n  → %s\n",
            n_events, n_predictors, n_predictors * 10,
            ifelse(n_events >= n_predictors * 10,
                   "Threshold met", "Below threshold — flag as limitation")))

# VIF
cat("\nVariance Inflation Factors (Logistic):\n")
vif_vals <- vif(glm_model)
print(round(vif_vals, 2))
cat(sprintf("  Max VIF: %.2f → %s\n", max(vif_vals),
            ifelse(max(vif_vals) < 5, "No multicollinearity concern", "Flag in paper")))

# Box-Tidwell test for linearity of log-odds
cat("\nBox-Tidwell test (linearity of log-odds):\n")
continuous_vars <- c("overall_index", "age", "duration_minutes")
bt_data <- predictors %>%
  select(all_of(c(continuous_vars, "stock_mkt_binary"))) %>%
  drop_na() %>%
  mutate(across(all_of(continuous_vars), ~ pmax(., 0.001))) %>%
  mutate(across(all_of(continuous_vars),
                list(int = ~ . * log(.)),
                .names = "{.col}_int"))

bt_formula <- reformulate(
  c(continuous_vars, paste0(continuous_vars, "_int")),
  response = "stock_mkt_binary"
)
bt_model <- glm(bt_formula, data = bt_data, family = binomial())
bt_summary <- coef(summary(bt_model))
bt_results <- bt_summary[grepl("_int", rownames(bt_summary)), , drop = FALSE]
print(round(bt_results, 4))
cat("  Non-significant interaction terms suggest linearity assumption holds.\n")
```

```{r warning=FALSE}
# LOGISTIC REGRESSION: STOCK MARKET ~ FINANCIAL LITERACY
glm_model2 <- glm(
  stock_mkt_binary ~ overall_index,
  data   = predictors,
  family = binomial(link = "logit")
)

summary(glm_model2)

or_lit <- data.frame(
  Odds_Ratio = exp(coef(glm_model2)),
  CI_Lower   = exp(confint(glm_model2)[, 1]),
  CI_Upper   = exp(confint(glm_model2)[, 2]),
  p_value    = coef(summary(glm_model2))[, 4]
)

print(round(or_lit, 3))
```

```{r warning=FALSE}
stargazer(
  lm_model, glm_model, glm_model2,
  type          = "html",
  out           = "regression_table.html",
  title         = "Table A1. Regression Results",
  column.labels = c("Linear Regression (OLS)",
                    "Logistic Regression (Full)",
                    "Logistic Regression (Literacy Only)"),
  model.numbers = FALSE,
  covariate.labels = c(
    "Age",
    "Survey duration (mins)",
    "Born in Canada",
    "Gender: Male",
    "Undergrad: Education & Child Studies",
    "Undergrad: Humanities & Languages",
    "Undergrad: Social Sciences",
    "Undergrad: STEM & Health Sciences",
    "Overall literacy index"
  ),
  add.lines = list(c("Significance levels", "* p<0.1", "** p<0.05", "*** p<0.01")),
  df = TRUE
)

save(lm_model, glm_model, glm_model2, 
     file = "../paper & figures/models.RData")
```

# Hypothesis test
```{r}
# One-sample proportion z-test
# H0: p = 0.33 (sample rate is equal to Canadian average)
# H1: p > 0.33 (sample rate is higher than Canadian average)

n <- 145       # sample size
x <- round(0.42 * n)  # number of investors (42%)
p_null <- 0.33       # Canadian national average

result <- prop.test(
  x  = x,
  n  = n,
  p  = p_null,
  alternative = "greater",  # one-tailed: testing if rate is higher
  correct   = FALSE         # no continuity correction, to match z-test
)

# Extract and report
z <- qnorm(result$p.value, lower.tail = FALSE)

cat("One-Sample Proportion Z-Test\n")
cat("-----------------------------\n")
cat(sprintf("Sample proportion:   %.2f\n", x/n))
cat(sprintf("Null proportion:     %.2f\n", p_null))
cat(sprintf("Z-statistic:         %.2f\n", z))
cat(sprintf("P-value:             %.3f\n", result$p.value))
cat(sprintf("\nConclusion: The sample participation rate (%.0f%%) is %s\n",
    x/n * 100,
    ifelse(result$p.value < 0.05,
           "statistically significantly higher than the Canadian average of 33% (p < 0.05).",
           "not statistically significantly higher than the Canadian average of 33%.")))
```