---
title: "Financial Literacy Analysis"
author: "Mariana Garcia Mejia"
output:
  html_document: default
  pdf_document: default
---

```{r message=FALSE, warning=FALSE}
# LOAD LIBRARIES
library(tidyverse)   # dplyr, tidyr, ggplot2
library(scales)      # percent formatting on axes
```

```{r load-data}
# LOAD DATASET
df <- read.csv("analysis_data.csv", fileEncoding = "latin1")
```

## 1. Exploratory Data Analysis

```{r scores-summary}
# Summary of the scores
indices <- c("advanced", "basic", "overall")
averages <- c(mean(df$advanced_index), mean(df$basic_index), mean(df$overall_index))
sd <- c(sd(df$advanced_index), sd(df$basic_index), sd(df$overall_index))

summary <- tibble::tibble(indices, averages, sd)
summary
```

```{r gender-pie}
# Gender distribution — pie chart
gender_counts <- df %>%
  count(gender) %>%
  filter(!is.na(gender))

ggplot(gender_counts, aes(x = "", y = n, fill = gender)) +
  geom_col(width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(n / sum(n) * 100), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  labs(title = "Gender Distribution", fill = NULL)
```

```{r summary-gender}
# Stock market participation by gender
summary_gender <- df %>%
  filter(gender %in% c("male", "female", "non binary")) %>%
  group_by(gender) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    n_no               = n_obs - n_yes,
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  )

print(summary_gender)
```

```{r summary-undergrad}
# Stock market participation by undergraduate category
summary_undergrad <- df %>%
  group_by(undergrad_category) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    n_no               = n_obs - n_yes,
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  )

print(summary_undergrad)
```

```{r summary-country}
# Stock market participation by country of birth
summary_country <- df %>%
  group_by(born_in_canada) %>%
  summarise(
    n_obs              = n(),
    n_yes              = sum(stock_mkt_participation_clean, na.rm = TRUE),
    participation_rate = round(mean(stock_mkt_participation_clean, na.rm = TRUE) * 100, 2)
  ) %>%
  mutate(born_in_canada = ifelse(born_in_canada == 1, "Born in Canada", "Born Outside Canada"))

print(summary_country)
```

```{r summary-scores-binary}
# Average literacy scores by stock market participation
summary_scores <- df %>%
  group_by(stock_mkt_participation_clean) %>%
  summarise(
    n_obs         = n(),
    avg_basic     = round(mean(basic_index,    na.rm = TRUE), 2),
    avg_advanced  = round(mean(advanced_index, na.rm = TRUE), 2),
    avg_overall   = round(mean(overall_index,  na.rm = TRUE), 2)
  ) %>%
  mutate(stock_mkt_participation_clean = ifelse(
    stock_mkt_participation_clean == 1,
    "Participates in Stock Market", "Does Not Participate / Refusal"
  ))

print(summary_scores)
```

# 2. Figures in the final paper

## Figure 1: Distribution of Index Scores
```{r figure1}
# Figure 1: Distribution of Index Scores
df %>%
  select(basic_index, advanced_index, overall_index) %>%
  pivot_longer(everything(), names_to = "Index", values_to = "Value") %>%
  mutate(Index = recode(Index,
                        "basic_index"    = "Basic Index",
                        "advanced_index" = "Advanced Index",
                        "overall_index"  = "Overall Index")) %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30, fill = "#0072B2", color = "white") +
  facet_wrap(~Index) +
  labs(
    title = "Distribution of Index Scores",
    x = "Value", y = "Frequency"
  ) +
  theme_minimal()
```

## Figure 2: Proportion of Correct Responses by Question Type
```{r figure2}
basic_questions    <- c("q1_basic_correct", "q2_basic_correct",
                        "q3_basic_correct", "q4_basic_correct")
advanced_questions <- c("q1_advanced_correct", "q2_advanced_correct", "q3_advanced_correct",
                        "q4_advanced_correct", "q5_advanced_correct")

basic_labels    <- c("Q1 — Numeracy", "Q2 — Inflation",
                     "Q3 — Time Value", "Q4 — Money Illusion")
advanced_labels <- c("Q1 — Stock Market Function", "Q2 — Stock Ownership",
                     "Q3 — Bond Ownership", "Q4 — Diversification",
                     "Q5 — Bond/Interest Rate")

basic_df <- data.frame(
  Question = basic_labels,
  Proportion = colMeans(df[, basic_questions], na.rm = TRUE),
  Type = "Basic"
)
advanced_df <- data.frame(
  Question = advanced_labels,
  Proportion = colMeans(df[, advanced_questions], na.rm = TRUE),
  Type = "Advanced"
)

all_q <- bind_rows(basic_df, advanced_df)

ggplot(all_q, aes(x = Question, y = Proportion)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~Type, scales = "free_x") +
  labs(
    title = "Proportion of Correct Responses by Question Type",
    x = NULL, y = "Proportion Correct"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Figure 3: Questions Breakdown


## Figure 4: Financial Literacy by Gender
```{r figure4}
df_gender <- df %>%
  filter(gender %in% c("male", "female")) %>%
  group_by(gender) %>%
  summarise(
    Basic    = mean(basic_index,    na.rm = TRUE),
    Advanced = mean(advanced_index, na.rm = TRUE),
    Overall  = mean(overall_index,  na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Basic, Advanced, Overall),
               names_to = "Index", values_to = "Score")

palette <- c("female" = "#FF7F0E", "male" = "#0072B2")

ggplot(df_gender, aes(x = Index, y = Score, fill = gender)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = palette, labels = c("Female", "Male")) +
  ylim(0, 1) +
  labs(
    title = "Financial Literacy by Gender",
    x = NULL, y = "Average Score", fill = NULL
  ) +
  theme_minimal()
```

## Figure 5: Stock Market Participation Pie Chart
```{r figure5}
stock_counts <- df %>%
  count(stock_mkt_participation) %>%
  filter(!is.na(stock_mkt_participation))

ggplot(stock_counts, aes(x = "", y = n, fill = stock_mkt_participation)) +
  geom_col(width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(round(n / sum(n) * 100), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  labs(
    title = "Do you own individual stocks, bonds, mutual funds or index funds?",
    fill = NULL
  )
```

## Figure 6: Stock Market Participation by Undergraduate Category
```{r}
# Figure 6: Stock Market Participation by Undergraduate Category
df %>%
  mutate(Participation = ifelse(stock_mkt_participation_clean == 1, "Yes", "No")) %>%
  ggplot(aes(x = undergrad_category, fill = Participation)) +
  geom_bar(position = "dodge") +
  labs(
    title  = "Stock Market Participation by Undergraduate Category",
    x      = "\nUndergraduate Category",
    y      = "Number of Respondents",
    fill   = "Stock Market Participation"
  ) +
  theme_minimal()
```

## Figure 7: Financial Literacy Scores by Stock Market Participation
```{r figure7}
means_long <- df %>%
  group_by(stock_mkt_participation_clean) %>%
  summarise(
    Basic    = mean(basic_index,    na.rm = TRUE),
    Advanced = mean(advanced_index, na.rm = TRUE),
    Overall  = mean(overall_index,  na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Basic, Advanced, Overall),
               names_to = "Index", values_to = "Score") %>%
  mutate(participation = ifelse(stock_mkt_participation_clean == 1, "Yes", "No / Refusal"))

ggplot(means_long, aes(x = participation, y = Score, fill = Index)) +
  geom_col(position = "dodge") +
  ylim(0, 0.8) +
  labs(
    title = "Financial Literacy Scores by Stock Market Participation",
    x = NULL, y = "Average Financial Literacy Score (%)", fill = NULL
  ) +
  theme_minimal()
```
